---
title: "Insomnia Simulation Study"
output:
  html_document: default
  word_document: default
  pdf_document: default
date: "2024-03-18"

---

```{r}
# Packages and their versions that are needed to replicate the simulation study
# results.

#install.packages("knitr")
#install.packages("truncnorm")
#install.packages("PearsonDS")
#install.packages("mice")
#install.packages("moments")
#install.packages("ggplot2")
#install.packages("dplyr")
#install.packages("ggpubr")
#install.packages("finalfit")
#install.packages("ggmice")
#install.packages("naniar")
#install.packages("lme4")
#install.packages("lmerTest")
#install.packages("car")
#install.packages("broom.mixed")

```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
# Removes the message related to the package version and information 

library(truncnorm)    # Simulates data from truncated normal distributions
library(PearsonDS)    # Simulates data from certain distributions
library(mice)         # Imputes data by using MICE
library(moments)      # Computes skewness and kurtosis
library(ggplot2)      # Advanced plotting
library(dplyr)        # Data handling and aggregations
library(ggpubr)       # Combines ggplots into one figure
library(finalfit)     # Analyses and visualises missingness
library(ggmice)       # Visualises missingness patterns
library(naniar)       # Performs Little's test for missingness (MAR vs. MCAR)
library(lme4)         # Fits linear mixed models
library(lmerTest)     # Computes p-values for the linear mixed models
library(car)          # Creates a qqplot to check the model assumptions for LMM
library(broom.mixed)  # Compute the pooled parameter estimates for LMM
```

```{r}
# Plotting colours
fig_colors <- c("#99c7e0", "#99d8c7")
```

# 1. Project Description

This project recreates and illustrates data analysis methods used in a pilot
study. Data was analysed to determine the influence of online cognitive
behavioural therapy for insomnia (CBT-I) on students suffering from insomnia.
The goal was to further analyse what types of problems the research setting
could encounter with the data. Due to data and client confidentiality, details
on the project description have been edited and the data for the analysis will
be fully simulated based on information from published articles to create a
representative showcase of the actual data analysis process that took place with
the real data.

In the project, the study recruited 98 students who suffered from insomnia and
randomly divided the students into a control group (no treatment) or the (CBT-I)
treatment group. The following were measured at the baseline:

-   Insomnia Severity Index (ISI) score to measure insomnia: the score ranges
    from 0 to 28; a minimum of 8 is needed to have an indication of insomnia.

-   Patient Health Questionnaire (PHQ-9) score to measure depression: the score
    ranges from 0 to 27; a minimum of 5 is needed to have an indication of mild
    symptoms of depression.

-   General Anxiety Disorder Assessment (GAD-7) to measure anxiety: the score
    ranges from 0 to 21 and a minimum score of 5 is needed to have an indication
    of mild symptoms.

The online CBT-I lasted for 5 weeks and after 9 weeks, ISI, PHQ-9 and GAD-7 was
again measured from all the participants in the treatment and in the control
group.

The following variables are considered the target variable:

-   ISI score

-   PHQ-9 score

-   GAD-7 score

The following variables are considered the independent variables:

-   Treatment with two levels: online CBT-I treatment group and control group

-   Time with two levels: pre-measurement (baseline) and post-measurements

The main research question of the project was: *What is the influence of the
treatment on students' levels of insomnia, depression and anxiety?* The further
goal was to later conduct a similar study, in which the participants would be
followed for a longer period of time. Therefore, the goal was also to determine,
based on the data, what types of challenges the research setting could expect to
encounter (e.g. participant dropouts) and how they could be taken into
consideration when planning the larger study.

# 2. Data Simulation

For the simulation, the mean and standard deviation for the participants' age,
ISI, PHQ-9 and GAD-7 score for the treatment and control group are based on the
results of a similar but much larger study by Freeman et al. (2017). There is no
prior information about the distribution for ISI in the population. However, as
the sample size is relatively large and the score should be bound between 8-28,
a truncated normal distribution was used to simulate baseline scores for all
participants. The same was also used to simulate the age for the participants.

Prior research indicates that the PHQ-9 and GAD-7 could be positively skewed in
the population (Akram et al., 2023, van der Zweerde et al., 2018). The data was
simulated from a Pearson distribution with the *PearsonDS()* library using
arbitrary skewness and kurtosis values to imitate the distributions.

The change in each score for post-results was simulated from a truncated normal
distribution with the mean change and post-results SD that was observed by
Freeman et al. (2017) for each group.

```{r}
set.seed(01012024)

# Source utility functions for the simulation
source("data_simulation_functions.R")

n <- 98  # Sample size for each measured time point (pre/post)

# Simulate a complete dataset with simulated missingness (MAR)
complete_data <- get_complete_sim_data(n)
data          <- get_data_with_missingness(complete_data)

```

```{r}
head(data)
```

# 3. Data Exploration

The typing of each variable is checked and correct levels/categories are added
for the discrete variables.

```{r}
# Check the class of each variable
sapply(data, class)

# Convert categorical variables to factors and add corresponding levels
data$Gender             <- as.factor(data$Gender)
levels(data$Gender)     <- c("Male", "Female")

data$Treatment          <- factor(data$Treatment)
levels(data$Treatment)  <- c("Control", "Treatment")

data$Time               <- as.factor(data$Time)
levels(data$Time)       <- c("Pre", "Post")

sapply(data, class)

```

The data is further divided into four subsets for later analysis and plots:
baseline results and post-measurements based on the treatment group.

```{r}
# Subset the dataset by time and treatment
data_pre            <- subset(data, Time == "Pre")
data_pre_control    <- subset(data_pre, Treatment == "Control")
data_pre_treatment  <- subset(data_pre, Treatment == "Treatment")

data_post           <- subset(data, Time == "Post")
data_post_control   <- subset(data_post, Treatment == "Control")
data_post_treatment <- subset(data_post, Treatment == "Treatment")

```

## 3.1 Univariate Analysis

```{r}
# Source the functions for the univariate analyses
source("univariate_analysis_functions.R")
```

### 3.1.1 Univariate Descriptive Statistics

In univariate analysis each variable is examined individually. First, the
descriptive statistics for the whole dataset were computed.

```{r}
get_descriptive_stats(data)
```

As 98 participants were recruited and measured twice, there should be a total of
196 observations. However, the number of observations is 181, indicating that 15
participants dropped out. In addition, *PHQ-9* is missing 13% (n=24) and *GAD-7*
14% (n=26) observations. The mean age of participants is 25.75 and we can see
that there are more females than males in the whole sample.

The descriptive statistics for the baseline measurements are the following:

```{r}
get_descriptive_stats(data_pre)
```

There are overall 98 participants in the baseline measurements with mean age of
25.93. *PHQ-9* is missing 14% of the observations (n=14) and *GAD-7* is missing
19% (n=19).

The descriptive statistics for the post-measurements are:

```{r}
get_descriptive_stats(data_post)
```

We can see that 15 people dropped out as the number of participants dropped to
83. Overall, *PHQ-9* and *GAD-7* seem to be missing values during
pre-measurements but there seems to be participants who dropped out, as the
number of observations dropped from 98 to 85.

Frequency tables are used to observe the proportion of participants across the
categorical variables.

```{r}
get_category_counts(data[, 3:5])
```

NOTE! As we have two measurements per participant, every count is doubled for
each participant (assuming they completed both measurements).

The frequency table grouped by all the categorical variables is the following:

```{r, message=FALSE, warning=FALSE}
# The warning about the change in the data structure (grouped variables as 
# attributes) is suppressed

output_table <- data                %>% 
  group_by(Gender, Time, Treatment) %>% 
  summarise(Count=n())
    
output_table
    
```

### 3.1.2 Distribution of the Numerical Variables

The distribution of the numerical variables will be examined by histograms and
boxplots.

#### 3.1.2.1 Histograms

```{r message=FALSE, warning=FALSE}
# Warnings about the missing data are suppressed

# The column indexes of the numerical variables
numerical_variable_columns_index <- c(2, 6, 7, 8)

# Loops each numerical column
for (i in numerical_variable_columns_index){
  x           <- data[, i]      # Variable
  x_var_name  <- names(data)[i] # Variable name
  
  # Histogram of the variable
  x_fig <- get_hist_grouped_ggplot(data        = data,
                                   x           = x,
                                   x_name      = x_var_name,
                                   group_var   = data$Treatment,
                                   main_title  = paste0("Histogram of ", x_var_name),
                                   x_title     = paste0(x_var_name))
  print(x_fig)

}
```

#### 3.1.2.2 Box plots

Next, the distribution of the numerical variables and the presence of outliers
is evaluated by analysing box plots.

```{r message=FALSE, warning=FALSE}
# Warnings about the missing data are suppressed

for (i in numerical_variable_columns_index){
  x          <- data[, i]      # Variable
  x_var_name <- names(data)[i] # Name of the variable

  # Box plot of the variable
  x_fig <- get_boxplot_grouped_ggplot(data       = data, 
                                      x          = x,
                                      group_var  = data$Treatment,
                                      x_title    = x_var_name,
                                      main_title = paste0("Boxplot of ", x_var_name))
  print(x_fig)

}
```

It can be seen that the variables *Age* and *PHQ-9* have a right-skewed
distribution. Some of the variables, especially *PHQ-9* seem to have outliers,
but as the values are within the allowed ranges, they can be left in the data as
they are.

#### 3.1.2.3 Test for Normality

As the figures have demonstrated that some of the variables show non-normal
distributions, the normality of each variable is further analysed by normality
tests. As each variable has duplicates of the same value, the normality is
assessed by using the D’Agostino test for skewness and the Anscombe-Glynn test
of kurtosis. The test is performed separately for the two treatment groups. The
tests compare the skewness and kurtosis of the observed data to the skewness and
kurtosis of a normal distribution.

The results for the two groups during the pre-measurements are the following:

```{r}
# Results for the control group
get_normality_test_results(data_pre_control, numerical_variable_columns_index) 
```

```{r}
# Results for the treatment group
get_normality_test_results(data_pre_treatment, numerical_variable_columns_index)

```

Next, the skewness and kurtosis test is performed separately for the
post-measurement results of the two groups.

```{r}
# Results for the control group
get_normality_test_results(data_post_control, numerical_variable_columns_index) 
```

```{r}
# Results for the control group
get_normality_test_results(data_post_treatment, numerical_variable_columns_index) 
```

For the pre-measurement the results indicate that *Age*, *ISI* and *GAD-7*
follow relatively normal distributions whereas *PHQ-9* has a right-skewed
distribution. For the post-measurements, *ISI* and *GAD-7* seem to follow a
normal distribution, whereas *Age* and *PHQ-9* have a right-skewed distribution.

#### 3.2 Frequency of Categorical Variables: Bar charts

```{r}
source("univariate_analysis_functions.R")
# Bar chart showing the number of females and males
barchart_gender_fig <- get_barchart_grouped_ggplot(data        = data,
                                                   x           = data$Gender,
                                                   x_labels    = c("Male", "Female"),
                                                   group_var   = data$Treatment,
                                                   main_title  = paste0("Barchart of the Number of Respondents by Gender"),
                                                   x_title     = "Gender")
print(barchart_gender_fig)

# Bar chart showing the number of participants in each treatment group
barchart_treatment_fig <- get_barchart_grouped_ggplot(data           = data,
                                                      x              = data$Time,
                                                      x_labels       = c("Pre", "Post"),
                                                      group_var      = data$Treatment,
                                                      legend_title   = "Treatment",
                                                      legend_labels  = c("Control", "Treatment"),
                                                      main_title     = paste0("Barchart of the Number of Respondents by Time"),
                                                      x_title        = "Time")
print(barchart_treatment_fig)

```

## 3.2 Bivariate Analysis

The relationship between multiple variables is also examined by plots and
correlation analysis.

```{r}
# Source the functions for the multivariate analyses
source("multivariate_analysis_functions.R")
```

### 3.2.1 Continuous Variables

The relationship between each numerical target variable is examined by analysing
the scatter plots and performing correlations tests.

#### 3.2.1.1 Scatter plots

First, the scatter plots between *ISI* and other numerical variables:

```{r message=FALSE, warning=FALSE}
# Warnings about the missing data are suppressed

# The column indexes for all numerical variables except ISI
numerical_variable_columns_index <- c(2, 7, 8)

for (i in numerical_variable_columns_index){
  y           <- data[, i]
  y_var_name  <- names(data)[i]
  

  fig <- get_scatterplot(data        = data,
                         x           = data$ISI,
                         y           = y,
                         x_title     = "ISI",
                         y_title     = y_var_name,
                         main_title  = paste0("Association between ISI and ", y_var_name))
  print(fig)
  
}
```

Next, the scatter plots between *PHQ-9* and other numerical variables:

```{r message=FALSE, warning=FALSE}
# Warnings about the missing data are suppressed

# The column indexes for all numerical variables except PHQ-9
numerical_variable_columns_index <- c(2, 6, 8)

for (i in numerical_variable_columns_index){
  y           <- data[, i]
  y_var_name  <- names(data)[i]
  

  fig <- get_scatterplot(data       = data,
                         x          = data$`PHQ-9`,
                         y          = y,
                         x_title    = "PHQ-9",
                         y_title    = y_var_name,
                         main_title = paste0("Association between PHQ-9 and ", y_var_name))
  print(fig)

}
```

Finally, the scatter plots between *GAD-7* and other numerical variables:

```{r message=FALSE, warning=FALSE}
# Warnings about the missing data are suppressed

# The column indexes for all numerical variables except GAD-7
numerical_variable_columns_index <- c(2, 6, 7)

for (i in numerical_variable_columns_index){
  y           <- data[, i]
  y_var_name  <- names(data)[i]
  

  fig <- get_scatterplot(data       = data,
                        x           = data$`GAD-7`,
                        y           = y,
                        x_title     = "GAD-7",
                        y_title     = y_var_name,
                        main_title  = paste0("Association between GAD and ", y_var_name))
  print(fig)

}
```

Most plots seem to indicate that there is no linear relationship between the
variables. This will be further examined with correlation analyses.

#### 3.2.1.2 Correlation Analysis

Next, the relationship between the variables that demonstrated a linear
relationship is examined separately for the pre- and post-results. If both
variables have a relatively normal distribution according to the normality test,
the Pearson correlation method is used. If one or both variables had a
non-normal (e.g. skewed) distribution, the Spearman correlation will be used
instead. The null hypothesis for both tests is that there is no correlation.

```{r message=FALSE, warning=FALSE}
# The warnings about not being able to compute an exact p-value with the Spearman method is suppressed

# Analysis for pre-measurements for the control group using
numerical_variable_columns_index <- c(6, 7, 8)

corr_results_pre_control <- get_corr_results_pre_control(data_pre_control,
                                                         numerical_variable_columns_index)
# Correlation analysis results for pre-control data
corr_results_pre_control
```

```{r message=FALSE, warning=FALSE}
# The warnings about not being able to compute an exact p-value with the Spearman method is suppressed

corr_results_pre_treatment <- get_corr_results_pre_treatment(data_pre_treatment,
                                                         numerical_variable_columns_index)
# Correlation analysis results for pre-treatment data
corr_results_pre_treatment

```

Based on the results, the only correlation that is significant (α = 0.05) is the
correlation between PHQ-9 and GAD-7 in the control group.

Next, the post-measurements are analysed by performing the correlation analysis
separately for each group.

```{r message=FALSE, warning=FALSE}
# The warnings about not being able to compute an exact p-value with the Spearman method is suppressed

corr_results_post_control <- get_corr_results_post_control(data_post_control,
                                                           numerical_variable_columns_index)

# Correlation analysis results for post-control data
corr_results_post_control

```

```{r message=FALSE, warning=FALSE}
# The warnings about not being able to compute an exact p-value with the Spearman method is suppressed

corr_results_post_treatment <- 
  get_corr_results_post_treatment(data_post_treatment,
                                  numerical_variable_columns_index)

# Correlation analysis results for post-treatment data
corr_results_post_treatment
```

Based on the results, we cannot reject the null hypothesis that the correlation
between any of the variables is 0.

### 3.2.2 Categorical Variables

The association between categorical variables is examined by bar charts,
analysing contingency tables and using chi-square test of independence.

#### 3.2.2.1 Bar chart

```{r}

get_barchart_grouped_grid_ggplot(data        = data,
                                 x           = data$Gender,
                                 x_labels    = c("Male", "Female"),
                                 group_var   = data$Treatment,
                                 main_title  = paste0("Barchart of the Number of Respondents by Gender"),
                                 x_title     = "Gender")
```

It can be seen that the ratio between females and males is quite equal in the
control group (both in pre and post measurements), whereas there are noticeably
less males than females in the treatment group. The same information can be
observed from the contingency tables.

#### 3.2.2.2 Contingency tables

```{r}
# Creates a contingency table by using a formula
contingency_table <- xtabs(~ Time + Treatment + Gender, data)

# Adds the total margins to the contingency table
ftable(addmargins(contingency_table, margin = 2:3, list(Total = sum)))
```

```{r}
# Creates a contingency table by using a formula
contingency_table <- xtabs(~ Treatment + Gender, data)

# Adds the total margins to the contingency table
ftable(addmargins(contingency_table, margin = 1:2, list(Total = sum)))
```

```{r}
# Creates a contingency table by using a formula
contingency_table <- xtabs(~ Time + Gender, data)

# Adds the total margins to the contingency table
ftable(addmargins(contingency_table, margin = 1:2, list(Total = sum)))
```

```{r}
# Creates a contingency table by using a formula
contingency_table <- xtabs(~ Time + Treatment, data)

# Adds the total margins to the contingency table
ftable(addmargins(contingency_table, margin = 1:2, list(Total = sum)))
```

#### 3.2.2.3 Chi-square test of independence

The chi-square test of independence is done to analyse if the categorical
variables are independent or not. The null hypothesis is that the variables are
independent.

```{r}
# Chi-square test of Independence between variables Gender and Treatment
chisq.test(data$Gender, data$Treatment)
```

```{r}
# Chi-square test of Independence between variables Gender and Time
chisq.test(data$Gender, data$Time)
```

```{r}
# Chi-square test of Independence between variables Treatment and Time
chisq.test(data$Treatment, data$Time)
```

The results indicate that all three variables are independent of each other.

### 3.2.3 Numerical and Categorical Variables

The relationship between the categorical and numerical variables (the impact of
gender on all the target variables) is visualised by box plots. Further tests
are not yet performed as they are the main focus of the statistical analysis.

```{r message=FALSE, warning=FALSE}
# Warnings about the missing data are suppressed

numerical_variable_columns_index <- c(2, 6, 7, 8)

for (i in numerical_variable_columns_index){
  x           <- data[, i]
  x_var_name  <- names(data)[i]
  

  fig <- get_boxplot_grouped_grid(
                        data          = data,
                        x             = x,
                        y             = data$Gender,
                        y_levels      = c("Male", "Female"),
                        group_var     = data$Gender,
                        legend_title  = "Gender",
                        legend_labels = c("Male", "Female"),
                        x_title       = x_var_name)
  print(fig)

}

```

We can observe that the largest differences in the distributions occur with
*PHQ-9* between males and females, espcially in the control group.

```{r message=FALSE, warning=FALSE}
# Warnings about the missing data are suppressed

numerical_variable_columns_index <- c(2, 6, 7, 8)

for (i in numerical_variable_columns_index){
  x           <- data[, i]
  x_var_name  <- names(data)[i]
  

  fig <- get_boxplot_grouped_grid(
                        data           = data,
                        y              = data$Treatment,
                        y_levels       = c("Treatment", "Control"),
                        x              = x,
                        group_var      = data$Treatment,
                        legend_title   = "Treatment",
                        legend_labels  = c("Control", "Treatment"),
                        x_title        = x_var_name,
                        grid_time_only = TRUE)
  print(fig)

}
```


